---
title: "Data Quality Project"
output: html_notebook
---

Load data and packages

```{r}
library(arrow)
library(tidyverse)
library(data.table)
library(readxl)
library(knitr)
library(icd)
library(readr)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(patchwork)
library(broom)

year <- "2020"
garbage <- read_csv("/Users/amymann/Documents/Data Quality Project/data_quality/cause-codes/gbd_garbage_codes_with_descr.csv")
path = "/Users/amymann/Documents/Data Quality Project/data/parquet/"
setwd(path)
mortality <- read_parquet("mort2020.parquet")
```

Calculate fraction of ‘garbage’ codes in the data, their GBD severity profiles, and the top garbage codes.
```{r}
clean_icd <- function(x) {
  x <- toupper(x)               
  x <- str_remove_all(x, "[^A-Z0-9\\.]")  
  str_trim(x)
}

mortality <- mortality %>%
  mutate(icd10 = clean_icd(ucod))

# Perform a single join by icd10, adding gbd_severity from garbage
mortality_flagged <- mortality %>%
  left_join(garbage %>% select(icd10, description, gbd_severity), by = "icd10")

# Compute the garbage fraction
garbage_fraction <- mean(!is.na(mortality_flagged$gbd_severity), na.rm = TRUE)

mortality_garbage <- mortality_flagged %>%
  filter(!is.na(gbd_severity))

# Count deaths by severity category
severity_counts <- mortality_garbage %>%
  count(gbd_severity, name = "deaths")

# Calculate the total number of garbage-coded deaths
total_garbage_deaths <- sum(severity_counts$deaths)

# Calculate share of each severity category
severity_profile <- severity_counts %>%
  mutate(share = deaths / total_garbage_deaths)

# Find the most common garbage codes
top_garbage <- mortality_garbage %>%    
  count(icd10, description, gbd_severity, sort = TRUE, name = "deaths") %>% 
  mutate(share = deaths / sum(deaths))

top_garbage_10 <- top_garbage %>% slice_head(n = 10)

garbage_fraction
severity_profile
top_garbage_10

write_csv(top_garbage_10, paste("top_garbage_10", year, ".csv", sep=""))

```

Demographic breakdown the proportion of garbage codes.
```{r}
clean_icd <- function(x) {
  x <- toupper(x)               
  x <- str_remove_all(x, "[^A-Z0-9\\.]")  
  str_trim(x)
}

mort <- mortality %>%
  mutate(icd10 = clean_icd(ucod))

mort <- mort |>            
  mutate(has_factor =
           rowSums(across(starts_with("record_"),
                          ~ .x != "" & !is.na(.x)>1))-1)

mortality_flagged <- mort %>%
  left_join(garbage %>% select(icd10, description, gbd_severity), by = "icd10")


DT <- as.data.table(mortality_flagged)           # convert once
DT[, garbage := !is.na(gbd_severity)]            # logical flag

garbage_by <- function(dt, col) {
  dt[, .(n_deaths = .N,
         garbage   = sum(garbage),
         prop_garbage = sum(garbage)/.N),
     by = col][order(-prop_garbage)]
}

by_sex      <- garbage_by(DT, "sex")
by_race     <- garbage_by(DT, "racer5")
by_age_grp  <- garbage_by(DT, "ager52")
by_educ     <- garbage_by(DT, "educ2003")
by_marstat  <- garbage_by(DT, "marstat")
by_weekday  <- garbage_by(DT, "weekday")
by_placdth  <- garbage_by(DT, "placdth")
by_has_secondary <- garbage_by(DT, "has_factor")
by_industry      <- garbage_by(DT, "Ind_23")

demog_detail <- rbindlist(list(
  by_sex     [, .(demog = "sex",      group = sex,      n_deaths, garbage, prop_garbage)],
  by_race    [, .(demog = "racer5",   group = racer5,   n_deaths, garbage, prop_garbage)],
  by_age_grp [, .(demog = "ager52",   group = ager52,   n_deaths, garbage, prop_garbage)],
  by_educ    [, .(demog = "education",group = educ2003, n_deaths, garbage, prop_garbage)],
  by_marstat [, .(demog = "marstat",  group = marstat,  n_deaths, garbage, prop_garbage)],
  by_weekday [, .(demog = "weekday",  group = weekday,  n_deaths, garbage, prop_garbage)],
  by_placdth [, .(demog = "placdth",  group = placdth,  n_deaths, garbage, prop_garbage)],
  by_has_secondary [, .(demog = "has_factor",  group = has_factor,  n_deaths, garbage, prop_garbage)],
  by_industry [, .(demog = "Ind_23",  group = Ind_23,  n_deaths, garbage, prop_garbage)]
))

# ────────────────────────────
# 3 · save or inspect
# ────────────────────────────
write_csv(demog_detail, "garbage_proportion_by_demographics.csv")

# ─────────────────────────────────────────────────────────────────────────
# 1 · lookup vector: ager52 → readable label
# ─────────────────────────────────────────────────────────────────────────
age52_lbl <- c(
  "Under 1 h","1–23 h", "1 day", "2 days", "3 days", "4 days", "5 days","6 days",
  "7–13 d", "14–20 d", "21–27 d","1 mo", "2 mo","3 mo","4 mo","5 mo","6 mo","7 mo", 
  "8 mo", "9 mo",  "10 mo", "11 mo", "1 y", "2 y","3 y", "4 y","5–9 y", "10–14 y",           
  "15–19 y", "20–24 y",  "25–29 y",  "30–34 y",  "35–39 y",
  "40–44 y", "45–49 y",  "50–54 y",  "55–59 y", "60–64 y",   "65–69 y",   "70–74 y",   "75–79 y", 
  "80–84 y",  "85–89 y", "90–94 y",  "95–99 y",                    
  "100–104 y", "105–109 y","110–114 y", "115–119 y", "120–124 y", "125–129 y", "130+ y"
)

# ─────────────────────────────────────────────────────────────────────────
# 2 · summarise proportion of garbage by ager52
# ─────────────────────────────────────────────────────────────────────────
by_age_grp_labeled <- by_age_grp %>%
  mutate(
    ager52 = as.integer(ager52),
    age_label = factor(age52_lbl[ager52],
                       levels = age52_lbl)
  )
# ────────────────────────────────────────────────────────────────
# 3 · plot: bar chart of garbage proportion by age-recode
# ────────────────────────────────────────────────────────────────
p <- ggplot(by_age_grp_labeled,
       aes(x = age_label, y = prop_garbage)) +
  geom_col(width = 0.9) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Proportion of garbage-coded deaths by age",
    x = "Age category",
    y = "Garbage codes (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1))
ggsave("garbage_by_age52.png", plot = p, width = 10, height = 6, dpi = 300)
```

Determine the number of different codes used. ***Note that this method might have a bug in it.
```{r}
library(readr); library(dplyr); library(stringr);library(tidyverse)
ihme <- read_csv("/Users/amymann/Documents/Data Quality Project/data_quality/cause-codes/ihme_cause_codes.csv")

clean_icd <- function(x) {
  x %>%
    str_to_upper() %>%                    # upper-case
    str_remove_all("[^A-Z0-9]") %>%    # keep letters, digits, dot
    str_trim()
}

mortality_clean <- mortality %>%
  mutate(icd10 = clean_icd(ucod)) 

ihme_clean <- ihme %>%
  mutate(icd10_code = clean_icd(icd10_code)) %>%     # assumes this column name
  filter(!is.na(icd10_code) & icd10_code != "") %>%  # drop empty rows
  distinct(icd10_code)                               # remove duplicates

observed_detail <- mortality_clean %>%
  distinct(icd10) %>%                         # unique codes in your data
  inner_join(ihme_clean, by = c("icd10" = "icd10_code")) %>%  # keep only valid codes
  nrow()

reference_detail <- nrow(ihme_clean)             
level_of_detail <- observed_detail / reference_detail
level_of_detail


```

Frequency of autopsies with demographic breakdown.
```{r}

# ── autopsy flag ────────────────────────────────────────────────────────────
mort <- mortality %>%
  mutate(
    autopsy_flag = recode(toupper(autopsy),   # normalise case
                          "Y" = "Yes",
                          "N" = "No",
                          .default = "Unknown")
  )

# ── detail-age  ───────────────────────────────
decode_age_detail <- function(detail_age){
  s     <- str_pad(detail_age, 4, pad = "0")
  unit  <- substr(s, 1, 1)              # 1=years,2=months,4=days,5=hrs,6=mins
  value <- as.numeric(substr(s, 2, 4))

  yrs <- dplyr::case_when(
    unit == "1" ~ value,
    unit == "2" ~ value / 12,
    unit == "4" ~ value / 365.25,
    unit == "5" ~ value / (24 * 365.25),
    unit == "6" ~ value / (60 * 24 * 365.25),
    TRUE        ~ NA_real_
  )
  round(yrs, 2)
}

age12_labels <- c(
  "01" = "<1 y",   "02" = "1-4 y",  "03" = "5-14 y",
  "04" = "15-24 y","05" = "25-34 y","06" = "35-44 y",
  "07" = "45-54 y","08" = "55-64 y","09" = "65-74 y",
  "10" = "75-84 y","11" = "85+ y",  "12" = "Age NS"
)

mort <- mort %>% mutate(
  age_years = decode_age_detail(age),
  # -- ager12 is numeric 1-12 in the file; pad with leading zero before recoding
  age_grp12 = fct_recode(sprintf("%02d", as.integer(ager12)), !!!age12_labels)
)

# ── education  ────────────────────────────────────────
educ03_map <- c(                         # 2003 revision :contentReference[oaicite:15]{index=15}:contentReference[oaicite:16]{index=16}
  "1"="≤8 grade","2"="9-12 no dip","3"="HS/GED",
  "4"="Some college","5"="Assoc deg","6"="Bach deg",
  "7"="Master","8"="Doctorate","9"="Edu unk"
)
educ89_map <- function(code){            # 1989 revision summary
  if (is.na(code) || code == "")            return("Edu unk")
  else if (code %in% sprintf("%02d", 1:8))  "≤8 grade"
  else if (code %in% c("09","10","11")) "Some HS"
  else if (code == "12")                "HS grad"
  else if (code %in% c("13","14","15")) "Some college"
  else if (code == "16")                "College grad"
  else if (code == "17")                "≥5 yrs college"
  else if (code == "00")                "None"
  else                                  "Edu unk"
}

mort <- mort %>%
  mutate(
    edu_final = case_when(
      educflag == 1 ~ recode(as.character(educ2003), !!!educ03_map),
      educflag == 0 ~ sapply(as.character(educ1989), educ89_map),
      TRUE          ~ "Edu missing"
    )
  )

# ──────── race ───────────────────────────────────

race5_map <- c(
  "1" = "White",
  "2" = "Black",
  "3" = "AI/AN",
  "4" = "Asian/PI",
  "0" = "Race unk"
)

mort <- mort %>% mutate(
  race_lbl = recode(as.character(racer5), !!!race5_map, .default = "Race unk"),
  hisp_lbl = recode(as.character(hspanicr),
                    "6" = "NH White",
                    "7" = "NH Black",
                    "8" = "NH Other",
                    "1" = "Mexican",
                    "2" = "PRican",
                    "3" = "Cuban",
                    "4" = "Cen/S Am",
                    "5" = "Other Hisp",
                    "9" = "Hisp unk",
                    .default = "Hisp unk")
)


# ── others ──────────────────────────────────────────────────────────
mort <- mort %>% mutate(
  sex_lbl  = recode(sex, "M" = "Male", "F" = "Female", .default = "Sex unk"),
  marital  = recode(marstat, "S"="Single","M"="Married","W"="Widowed",
                              "D"="Divorced","U"="MS unk", .default="MS unk"),
  inj_work = recode(toupper(injwork), "Y"="Yes","N"="No", .default="Unknown"),
  cause113 = ucr113
)

# ── set dimensions ──────────────────────────────────────
dims <- c("sex_lbl", "age_grp12", "race_lbl", "hisp_lbl",
          "edu_final", "marital", "inj_work", "cause113")

make_table <- function(df, var){
  df %>% 
    group_by(.data[[var]], autopsy_flag) %>% 
    summarise(deaths = n(), .groups = "drop_last") %>% 
    mutate(
      prop     = deaths / sum(deaths),
      variable = var,
      level    = as.character(.data[[var]])  
    ) %>% 
    ungroup() %>% 
    select(variable, level, autopsy_flag, deaths, prop)
}

autopsy_breakdown <- bind_rows(lapply(dims, make_table, df = mort))

library(readr)
lookup113 <- read_csv("cause_code_135.csv", show_col_types = FALSE)

autopsy_breakdown <- autopsy_breakdown %>%
  left_join(lookup113, by = c("level" = "code"))

# ── summary by sex ─────────────────────────────────────────────
sex_counts <- mort %>%
  filter(autopsy_flag %in% c("Yes", "No")) %>%  
  group_by(sex_lbl) %>%
  summarise(
    autopsy_yes = sum(autopsy_flag == "Yes"),
    total       = n(),
    .groups = "drop"
  )

sex_stat <- sex_counts %>%
  rowwise() %>%
  mutate(
    prop = autopsy_yes / total,
  ) %>%

print(sex_stat)

write_csv(autopsy_breakdown,
          "autopsy_breakdown_demographics_2016.csv")  
print(head(autopsy_breakdown, 20))
```

Is the male/female difference in autopsy simply a result of men dying younger? The rest of it could be the way men are dying? 
```{r}
reg_dat <- mort %>%                               # keep only Yes/No rows
  filter(autopsy_flag %in% c("Yes", "No")) %>%
  mutate(
    autopsy_bin = as.integer(autopsy_flag == "Yes")  # 1 = autopsy performed
  )

# model 1: sex only
m1 <- lm(autopsy_bin ~ sex_lbl, data = reg_dat)

# model 2: add age (continuous, in years)
m2 <- lm(autopsy_bin ~ sex_lbl + age_years, data = reg_dat)

m3 <- lm(autopsy_bin ~  age_years, data = reg_dat)

# quick summaries
summary(m1)          # baseline difference ⟶ coefficient for sex_lblMale
summary(m2)          # see how that coefficient shrinks once age is added
summary(m3)
anova(m3, m2)

# if you want a tidy comparison table:
bind_rows(
  tidy(m1) %>% mutate(model = "Sex only"),
  tidy(m2) %>% mutate(model = "Sex + age"),
  tidy(m3) %>% mutate(model = "Age only")
)
```

What is the effect of marital status on autopsy after controlling for age?
```{r}
reg_dat <- mortality %>%
  filter(autopsy %in% c("Y", "N")) %>%
  mutate(
    autopsy_bin = as.integer(autopsy == "Y")
  )


# model 1: sex only
m1 <- lm(autopsy_bin ~ marstat, data = reg_dat)

# model 2: add age (continuous, in years)
m2 <- lm(autopsy_bin ~ marstat + ager52, data = reg_dat)

# quick summaries
summary(m1)          # baseline difference ⟶ coefficient for sex_lblMale
summary(m2)          # see how that coefficient shrinks once age is added

# if you want a tidy comparison table:
bind_rows(
  tidy(m1) %>% mutate(model = "Marital status only"),
  tidy(m2) %>% mutate(model = "Marital status + age"),
)
```

Make plots for autopsy data!
```{r}

autopsy_breakdown <- read_csv(
  "/Users/amymann/Documents/Data Quality Project/data_quality/autopsy_breakdown_demographics_2016.csv",
  show_col_types = FALSE
)

age12_labels <- setNames(
  c("<1", "1–4", "5–14", "15–24", "25–34", "35–44",
    "45–54", "55–64", "65–74", "75–84", "85+", "NS"),
  sprintf("%02d", 1:12)
)

# ── 3. Helper to build one stacked-bar plot ───────────────────────────────────
plot_autopsy_prop <- function(df, var){
  plot_df <- df %>%
    filter(variable == var, autopsy_flag != "Unknown") %>%
    mutate(level = case_when(
      var == "age_grp12"              ~ factor(level,
                                               levels = names(age12_labels),
                                               labels = age12_labels),
      var %in% c("race_lbl", "hisp_lbl") ~ fct_relevel(level),  # keep file order
      TRUE                              ~ fct_infreq(level)     # order by freq
    ))

  ggplot(plot_df, aes(level, prop, fill = autopsy_flag)) +
    geom_col(width = 0.9, position = "stack") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_brewer(palette = "Set1") +
    labs(x = NULL, y = "Proportion of deaths",
         fill = "Autopsy performed",
         title = paste("Autopsy frequency by", var)) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title  = element_text(size = 12, face = "bold"))
}

# ── pick variables for grid  ──────────────────
grid_vars <- c("sex_lbl", "age_grp12", "race_lbl", "hisp_lbl",
               "edu_final", "marital", "inj_work")

# ── make grid plot  ──────────────────────────────────────────────────
grid_plot <- wrap_plots(
  lapply(grid_vars, \(v) plot_autopsy_prop(autopsy_breakdown, v)),
  ncol = 2
) +
  plot_annotation(
    title = "Autopsy proportions across demographic groups (US deaths, 2016)",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

print(grid_plot)
ggsave("autopsy_demographics_grid_2016.png",
       grid_plot, width = 14, height = 10, dpi = 300)

# ── make 113 cause groups plot ───────────────────────────
cause_plot <- autopsy_breakdown %>%
  filter(variable == "cause113",
         autopsy_flag != "Unknown",
         as.numeric(level) >= 1,
         as.numeric(level) <= 113) %>%
  mutate(desc = fct_reorder(level, prop)) %>%   # use plain-English label
  ggplot(aes(x = desc, y = prop, fill = autopsy_flag)) +
    geom_col(position = "stack", width = 0.9) +
    scale_y_continuous(labels = percent_format()) +
    scale_fill_manual(values = c("Yes"="#377eb8",
                                 "No" ="#e41a1c")) +
    labs(x = "Underlying-cause (113-group recode)",
         y = "Proportion of deaths",
         fill = "Autopsy",
         title = "Autopsy frequency by underlying cause, 2016") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4),
          plot.title  = element_text(size = 13, face = "bold"))

print(cause_plot)
ggsave("autopsy_by_cause113_2016.png",
       cause_plot, width = 16, height = 6, dpi = 300)

```

Average number of secondary factors listed 
```{r}
demo_vars <- c("race", "racer5", "educ2003", "sex", "ager52", "marstat", "ucr39")
mortality <- mortality %>% mutate(ranum = as.numeric(ranum)-1)

extra_factor_summary <- bind_rows(
  lapply(demo_vars, function(v)
    mortality %>% 
      group_by(.data[[v]], .drop = FALSE) %>% 
      summarise(
        n_deaths   = n(),
        mean_extra = mean(ranum-1, na.rm = TRUE),
        .groups    = "drop"
      ) %>% 
      transmute(
        variable   = v,
        level      = as.character(.data[[v]]),    # ensure character
        n_deaths,
        mean_extra
      )
  )
)


print(extra_factor_summary)
write_csv(extra_factor_summary, "avg_secondary_factors_by_demo.csv")

top10_ucr39 <- mort %>%
  count(ucr39, sort = TRUE, name = "n_deaths") %>%
  slice_head(n = 10) %>%
  left_join(
    mortality %>% group_by(ucr39) %>%
      summarise(mean_extra = mean(ranum-1, na.rm = TRUE),
                .groups = "drop"),
    by = "ucr39"
  )
print(top10_ucr39)


hist_plot <- ggplot(mortality, aes(ranum)) +
  geom_histogram(binwidth = 1, boundary = -0.5, fill = "#2e86c1") +
  scale_x_continuous(breaks = 0:20) +
  labs(x = "Number of record-axis conditions",
       y = "Deaths",
       title = "Distribution of additional conditions per death") +
  theme_minimal(base_size = 11)

ggsave("hist_extra_factors.png", hist_plot, width = 7, height = 4, dpi = 300)

age_var <- if ("age52" %in% names(mortality)) "age52" else "ager52"
age_means <- mortality %>%
  filter(!is.na(.data[[age_var]])) %>%
  group_by(ager52 = .data[[age_var]]) %>%
  summarise(
    n_deaths   = n(),
    mean_extra = -mean(ranum, na.rm = TRUE),
    .groups    = "drop"
  )

age_plot <- ggplot(age_means,
                   aes(factor(.data[[age_var]]), mean_extra)) +
  geom_col(fill = "#e41a1c") +
  scale_y_continuous(labels = number_format(accuracy = 0.1)) +
  labs(x = "Age category (Recode 52)",
       y = "Mean number of record-axis conditions",
       title = "Average additional conditions by age group") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave("mean_extra_factors_by_age.png", age_plot, width = 8, height = 4, dpi = 300)

```

Frequency of 0 secondary factors listed 
```{r}
demo_vars <- c("racer3", "racer5", "educ2003", "sex", "ager52", "marstat")


# ── 1. proportion with *zero* secondary factors, by demographic ───────────────
no_sec_summary <- bind_rows(
  lapply(demo_vars, function(v)
    mortality %>% 
      group_by(.data[[v]], .drop = FALSE) %>% 
      summarise(
        n_deaths   = n(),
        prop_zero  = mean(ranum == 0, na.rm = TRUE),
        .groups    = "drop"
      ) %>% 
      transmute(
        variable = v,
        level    = as.character(.data[[v]]),
        n_deaths,
        prop_zero
      )
  )
)

write_csv(no_sec_summary, "prop_no_secondary_by_demo.csv")
print(no_sec_summary)

# ── 2. top-10 UCR-39 causes – prop with no secondary factor ───────────────────
top10_ucr39_none <- mort %>%
  count(ucr39, sort = TRUE, name = "n_deaths") %>%
  slice_head(n = 10) %>%
  left_join(
    mort %>% group_by(ucr39) %>%
      summarise(prop_zero = mean(ranum == 0, na.rm = TRUE),
                .groups   = "drop"),
    by = "ucr39"
  )

write_csv(top10_ucr39_none, "prop_no_secondary_top10_ucr39.csv")
print(top10_ucr39_none)

# ── 3. bar chart of prop-zero by age52 (or ager52) ────────────────────────────
age_var <- if ("age52" %in% names(mort)) "age52" else "ager52"

age_prop_zero <- mort %>%
  group_by(.data[[age_var]]) %>%
  summarise(prop_zero = mean(extra_factor_count == 0, na.rm = TRUE),
            .groups   = "drop")

ggplot(age_prop_zero, aes(factor(.data[[age_var]]), prop_zero)) +
  geom_col(fill = "#e41a1c") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Age category (Recode 52)",
       y = "Share of deaths with NO secondary factor",
       title = "Percentage with no record-axis conditions by age group") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave("prop_no_secondary_by_age.png", width = 8, height = 4, dpi = 300)

```

Can sex/marital differences in frequency of 0 secondary factors listed be explained by cause of death and/or age of death?
```{r}
mort <- mortality %>% 
  mutate(
    has_extra_factor = as.numeric(ranum) > 1)

mort <- mort %>%                    
  mutate(
    ucr39 = factor(ucr39, levels = 1:39) 
  )

m1 <- lm(has_extra_factor ~ marstat + as.numeric(ager52), data = mort)


summary(m1)          
```

How correlated are the various indices (no secondary factors, average number of secondary factors, proportion of garbage codes, and whether or not their is an autopsy)?
```{r}
library(dplyr)
library(stringr)
library(tidyr)

# ----------------------------------------------------------------
# 1. Helper: clean ICD10 codes
# ----------------------------------------------------------------
clean_icd <- function(x) {
  x <- toupper(x)
  x <- str_remove_all(x, "[^A-Z0-9\\.]")
  str_trim(x)
}

# ----------------------------------------------------------------
# 2. Correlation function
# ----------------------------------------------------------------
calculate_dqi_correlations <- function(data,
                                       vars   = c("has_factor",
                                                  "avg_secondary_factor",
                                                  "prop_garbage_codes",
                                                  "autopsy_flag_num"),
                                       method = "spearman") {
  if (!all(vars %in% names(data))) {
    stop("One or more of the specified variables are missing in `data`.")
  }
  d <- data[vars]

  # Coerce logical / 2-level factor → 0/1 numeric
  to_numeric <- function(x) {
    if (is.logical(x)) return(as.numeric(x))
    if (is.factor(x) && nlevels(x) == 2) {
      return(as.numeric(as.integer(x) - 1))
    }
    return(x)
  }
  d[] <- lapply(d, to_numeric)

  # 1) Correlation matrix
  cor_mat <- cor(d, use = "pairwise.complete.obs", method = method)

  # 2) Tidy table of pairwise rho + p-value + n
  combs <- combn(vars, 2, simplify = FALSE)
  pair_stats <- lapply(combs, function(v) {
    x <- d[[v[1]]]]; y <- d[[v[2]]]]
    idx <- complete.cases(x, y)
    test <- cor.test(x[idx], y[idx], method = method, exact = FALSE)
    data.frame(
      var1    = v[1],
      var2    = v[2],
      rho     = unname(test$estimate),
      p_value = test$p.value,
      n       = sum(idx),
      stringsAsFactors = FALSE
    )
  })

  list(
    correlation_matrix = cor_mat,
    pairwise_table     = bind_rows(pair_stats)
  )
}

# ----------------------------------------------------------------
# 3. Build your data-quality indices on `mort`
# ----------------------------------------------------------------
mort <- mortality %>%
  # normalize ICD10
  mutate(icd10 = clean_icd(ucod)) %>%

  # 3a) no secondary factors flag (0=yes none; 1=has ≥1)
  mutate(
    has_factor = 1L - as.integer(
      rowSums(across(starts_with("record_"),
                     ~ !is.na(.x) & .x != "")) > 1
    )
  ) %>%

  # 3b) average number of secondary factors
  mutate(
    avg_secondary_factor =
      rowSums(across(starts_with("record_"),
                     ~ !is.na(.x) & .x != "")) - 1L
  ) %>%

  # 3c) autopsy flag → Yes/No → numeric 0/1
  mutate(
    autopsy_flag = recode(toupper(autopsy),
                          "Y" = "Yes", "N" = "No",
                          .default = "Unknown"),
    autopsy_flag_num = case_when(
      autopsy_flag == "Yes" ~ 1L,
      autopsy_flag == "No"  ~ 0L,
      TRUE                  ~ NA_integer_
    )
  )

# 3d) garbage-code flag via %in% (no join explosion)
garb_icds <- garbage %>%
  pull(icd10) %>%
  clean_icd() %>%
  unique()

mort <- mort %>%
  mutate(
    garbage_flag      = as.integer(icd10 %in% garb_icds),
    prop_garbage_codes = garbage_flag  # 0/1 per record
  )

# ----------------------------------------------------------------
# 4. Run correlations
# ----------------------------------------------------------------
results <- calculate_dqi_correlations(
  mort,
  vars = c("has_factor",
           "avg_secondary_factor",
           "prop_garbage_codes",
           "autopsy_flag_num")
)

# View results
results$correlation_matrix
results$pairwise_table

```

Occupation data
```{r}

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

