---
title: "Data Quality Project"
output: html_notebook
---

Load data and packages

```{r}
library(arrow)
library(tidyverse)
library(data.table)
library(readxl)
library(knitr)
library(icd)
library(readr)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(patchwork)
library(broom)

garbage <- read_csv("/Users/amymann/Documents/Data Quality Project/data_quality/cause-codes/gbd_garbage_codes_with_descr.csv")
path = "/Users/amymann/Documents/Data Quality Project/data/"
setwd(path)
mortality <- read_parquet("mort2016.parquet")

```

Model completeness which is the Reg CDR / True CDR (since we don't have access to True CDR). 
This is assuming that the percentage of under five deaths (true) is more accurate then the data registry. 

Predicted completeness usually within 3-5% of mean (R^2 = .85)
```{r}
RegCDR = nrow(mortality)/323071755 *1000.
sixtyfiveplus = 49200000/323071755 
Year = 2016.
gamma = 0.    # IDK what gamma is
births2016 <- 3945875                       # NCHS final birth count
under5 <- sum(
  with(mortality, {
    age4  <- sprintf("%04d", as.integer(age))      # keep leading zeros
    unit  <- substr(age4, 1, 1)
    value <- as.integer(substr(age4, 2, 4))

    # under-5 if: (unit = years & value < 5)  OR  (unit = months/days/hours/minutes)
    (unit == "1" & value < 5) | unit %in% c("2", "4", "5", "6")
  }),
  na.rm = TRUE
)
vr5q0      <- under5 / births2016           # probability of dying <5
fiveq0     <- 7 / 1000                      # IGME comparator (probability)
Compl_5q_0 <- vr5q0 / fiveq0                # should be ≈ 1 for the US
Compl_5q_0
logit_completeness = (RegCDR^2 * -0.0187471)+(RegCDR * 0.6125569)+(sixtyfiveplus * -12.58245)+(log(fiveq0) * -1.134923)+ (Compl_5q_0 * 2.319505)+(Year*-0.0184299)+31.40303+ gamma
completeness = exp(logit_completeness)/(1 + exp(logit_completeness))
completeness
```

Calculate fraction of ‘garbage’ codes in the data, their GBD severity profiles, and the top garbage codes.
```{r}
clean_icd <- function(x) {
  x <- toupper(x)               
  x <- str_remove_all(x, "[^A-Z0-9\\.]")  
  str_trim(x)
}

mortality <- mortality %>%
  mutate(icd10 = clean_icd(ucod))

# Perform a single join by icd10, adding gbd_severity from garbage
mortality_flagged <- mortality %>%
  left_join(garbage %>% select(icd10, description, gbd_severity), by = "icd10")

# Compute the garbage fraction
garbage_fraction <- mean(!is.na(mortality_flagged$gbd_severity), na.rm = TRUE)

mortality_garbage <- mortality_flagged %>%
  filter(!is.na(gbd_severity))

# Count deaths by severity category
severity_counts <- mortality_garbage %>%
  count(gbd_severity, name = "deaths")

# Calculate the total number of garbage-coded deaths
total_garbage_deaths <- sum(severity_counts$deaths)

# Calculate share of each severity category
severity_profile <- severity_counts %>%
  mutate(share = deaths / total_garbage_deaths)

# Find the most common garbage codes
top_garbage <- mortality_garbage %>%    
  count(icd10, description, gbd_severity, sort = TRUE, name = "deaths") %>% 
  mutate(share = deaths / sum(deaths))

top_garbage_10 <- top_garbage %>% slice_head(n = 10)

garbage_fraction
severity_profile
top_garbage_10

write_csv(top_garbage_10, "top_garbage_10.csv")

```

Demographic breakdown the proportion of garbage codes.
```{r}
clean_icd <- function(x) {
  x <- toupper(x)               
  x <- str_remove_all(x, "[^A-Z0-9\\.]")  
  str_trim(x)
}

mort <- mortality %>%
  mutate(icd10 = clean_icd(ucod))

mort <- mort |>            
  mutate(has_factor =
           rowSums(across(starts_with("record_"),
                          ~ .x != "" & !is.na(.x)>1))-1)

mortality_flagged <- mort %>%
  left_join(garbage %>% select(icd10, description, gbd_severity), by = "icd10")


DT <- as.data.table(mortality_flagged)           # convert once
DT[, garbage := !is.na(gbd_severity)]            # logical flag

garbage_by <- function(dt, col) {
  dt[, .(n_deaths = .N,
         garbage   = sum(garbage),
         prop_garbage = sum(garbage)/.N),
     by = col][order(-prop_garbage)]
}

by_sex      <- garbage_by(DT, "sex")
by_race     <- garbage_by(DT, "racer5")
by_age_grp  <- garbage_by(DT, "ager52")
by_educ     <- garbage_by(DT, "educ2003")
by_marstat  <- garbage_by(DT, "marstat")
by_weekday  <- garbage_by(DT, "weekday")
by_placdth  <- garbage_by(DT, "placdth")
by_has_secondary <- garbage_by(DT, "has_secondary")

demog_detail <- rbindlist(list(
  by_sex     [, .(demog = "sex",      group = sex,      n_deaths, garbage, prop_garbage)],
  by_race    [, .(demog = "racer5",   group = racer5,   n_deaths, garbage, prop_garbage)],
  by_age_grp [, .(demog = "ager52",   group = ager52,   n_deaths, garbage, prop_garbage)],
  by_educ    [, .(demog = "education",group = educ2003, n_deaths, garbage, prop_garbage)],
  by_marstat [, .(demog = "marstat",  group = marstat,  n_deaths, garbage, prop_garbage)],
  by_weekday [, .(demog = "weekday",  group = weekday,  n_deaths, garbage, prop_garbage)],
  by_placdth [, .(demog = "placdth",  group = placdth,  n_deaths, garbage, prop_garbage)],
  by_has_secondary [, .(demog = "has_secondary",  group = has_secondary,  n_deaths, garbage, prop_garbage)]
))

# ────────────────────────────
# 3 · save or inspect
# ────────────────────────────
write_csv(demog_detail, "garbage_proportion_by_demographics.csv")

# ─────────────────────────────────────────────────────────────────────────
# 1 · lookup vector: ager52 → readable label
# ─────────────────────────────────────────────────────────────────────────
age52_lbl <- c(
  "Under 1 h","1–23 h", "1 day", "2 days", "3 days", "4 days", "5 days","6 days",
  "7–13 d", "14–20 d", "21–27 d","1 mo", "2 mo","3 mo","4 mo","5 mo","6 mo","7 mo", 
  "8 mo", "9 mo",  "10 mo", "11 mo", "1 y", "2 y","3 y", "4 y","5–9 y", "10–14 y",           
  "15–19 y", "20–24 y",  "25–29 y",  "30–34 y",  "35–39 y",
  "40–44 y", "45–49 y",  "50–54 y",  "55–59 y", "60–64 y",   "65–69 y",   "70–74 y",   "75–79 y", 
  "80–84 y",  "85–89 y", "90–94 y",  "95–99 y",                    
  "100–104 y", "105–109 y","110–114 y", "115–119 y", "120–124 y", "125–129 y", "130+ y"
)

# ─────────────────────────────────────────────────────────────────────────
# 2 · summarise proportion of garbage by ager52
# ─────────────────────────────────────────────────────────────────────────
by_age_grp_labeled <- by_age_grp %>%
  mutate(
    age_label = factor(age52_lbl[ager52],
                       levels = age52_lbl)   # keep age order on x-axis
  )

# ────────────────────────────────────────────────────────────────
# 3 · plot: bar chart of garbage proportion by age-recode
# ────────────────────────────────────────────────────────────────
p <- ggplot(by_age_grp_labeled,
       aes(x = age_label, y = prop_garbage)) +
  geom_col(width = 0.9) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Proportion of garbage-coded deaths by age",
    x = "Age category",
    y = "Garbage codes (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1))
ggsave("garbage_by_age52.png", plot = p, width = 10, height = 6, dpi = 300)
```

Determine the number of different codes used. ***Note that this method might have a bug in it.
```{r}
library(readr); library(dplyr); library(stringr);library(tidyverse)
ihme <- read_csv("/Users/amymann/Documents/Data Quality Project/data_quality/ihme_cause_codes.csv")

clean_icd <- function(x) {
  x %>%
    str_to_upper() %>%                    # upper-case
    str_remove_all("[^A-Z0-9]") %>%    # keep letters, digits, dot
    str_trim()
}

mortality_clean <- mortality %>%
  mutate(icd10 = clean_icd(ucod)) 

ihme_clean <- ihme %>%
  mutate(icd10_code = clean_icd(icd10_code)) %>%     # assumes this column name
  filter(!is.na(icd10_code) & icd10_code != "") %>%  # drop empty rows
  distinct(icd10_code)                               # remove duplicates

observed_detail <- mortality_clean %>%
  distinct(icd10) %>%                         # unique codes in your data
  inner_join(ihme_clean, by = c("icd10" = "icd10_code")) %>%  # keep only valid codes
  nrow()

reference_detail <- nrow(ihme_clean)             
level_of_detail <- observed_detail / reference_detail
level_of_detail


```

Frequency of autopsies with demographic breakdown.
```{r}

# ── autopsy flag ────────────────────────────────────────────────────────────
mort <- mortality %>%
  mutate(
    autopsy_flag = recode(toupper(autopsy),   # normalise case
                          "Y" = "Yes",
                          "N" = "No",
                          .default = "Unknown")
  )

# ── detail-age  ───────────────────────────────
decode_age_detail <- function(detail_age){
  s     <- str_pad(detail_age, 4, pad = "0")
  unit  <- substr(s, 1, 1)              # 1=years,2=months,4=days,5=hrs,6=mins
  value <- as.numeric(substr(s, 2, 4))

  yrs <- dplyr::case_when(
    unit == "1" ~ value,
    unit == "2" ~ value / 12,
    unit == "4" ~ value / 365.25,
    unit == "5" ~ value / (24 * 365.25),
    unit == "6" ~ value / (60 * 24 * 365.25),
    TRUE        ~ NA_real_
  )
  round(yrs, 2)
}

age12_labels <- c(
  "01" = "<1 y",   "02" = "1-4 y",  "03" = "5-14 y",
  "04" = "15-24 y","05" = "25-34 y","06" = "35-44 y",
  "07" = "45-54 y","08" = "55-64 y","09" = "65-74 y",
  "10" = "75-84 y","11" = "85+ y",  "12" = "Age NS"
)

mort <- mort %>% mutate(
  age_years = decode_age_detail(age),
  # -- ager12 is numeric 1-12 in the file; pad with leading zero before recoding
  age_grp12 = fct_recode(sprintf("%02d", as.integer(ager12)), !!!age12_labels)
)

# ── education  ────────────────────────────────────────
educ03_map <- c(                         # 2003 revision :contentReference[oaicite:15]{index=15}:contentReference[oaicite:16]{index=16}
  "1"="≤8 grade","2"="9-12 no dip","3"="HS/GED",
  "4"="Some college","5"="Assoc deg","6"="Bach deg",
  "7"="Master","8"="Doctorate","9"="Edu unk"
)
educ89_map <- function(code){            # 1989 revision summary
  if (is.na(code) || code == "")            return("Edu unk")
  else if (code %in% sprintf("%02d", 1:8))  "≤8 grade"
  else if (code %in% c("09","10","11")) "Some HS"
  else if (code == "12")                "HS grad"
  else if (code %in% c("13","14","15")) "Some college"
  else if (code == "16")                "College grad"
  else if (code == "17")                "≥5 yrs college"
  else if (code == "00")                "None"
  else                                  "Edu unk"
}

mort <- mort %>%
  mutate(
    edu_final = case_when(
      educflag == 1 ~ recode(as.character(educ2003), !!!educ03_map),
      educflag == 0 ~ sapply(as.character(educ1989), educ89_map),
      TRUE          ~ "Edu missing"
    )
  )

# ──────── race ───────────────────────────────────

race5_map <- c(
  "1" = "White",
  "2" = "Black",
  "3" = "AI/AN",
  "4" = "Asian/PI",
  "0" = "Race unk"
)

mort <- mort %>% mutate(
  race_lbl = recode(as.character(racer5), !!!race5_map, .default = "Race unk"),
  hisp_lbl = recode(as.character(hspanicr),
                    "6" = "NH White",
                    "7" = "NH Black",
                    "8" = "NH Other",
                    "1" = "Mexican",
                    "2" = "PRican",
                    "3" = "Cuban",
                    "4" = "Cen/S Am",
                    "5" = "Other Hisp",
                    "9" = "Hisp unk",
                    .default = "Hisp unk")
)


# ── others ──────────────────────────────────────────────────────────
mort <- mort %>% mutate(
  sex_lbl  = recode(sex, "M" = "Male", "F" = "Female", .default = "Sex unk"),
  marital  = recode(marstat, "S"="Single","M"="Married","W"="Widowed",
                              "D"="Divorced","U"="MS unk", .default="MS unk"),
  inj_work = recode(toupper(injwork), "Y"="Yes","N"="No", .default="Unknown"),
  cause113 = ucr113
)

# ── set dimensions ──────────────────────────────────────
dims <- c("sex_lbl", "age_grp12", "race_lbl", "hisp_lbl",
          "edu_final", "marital", "inj_work", "cause113")

make_table <- function(df, var){
  df %>% 
    group_by(.data[[var]], autopsy_flag) %>% 
    summarise(deaths = n(), .groups = "drop_last") %>% 
    mutate(
      prop     = deaths / sum(deaths),
      variable = var,
      level    = as.character(.data[[var]])  
    ) %>% 
    ungroup() %>% 
    select(variable, level, autopsy_flag, deaths, prop)
}

autopsy_breakdown <- bind_rows(lapply(dims, make_table, df = mort))

library(readr)
lookup113 <- read_csv("cause_code_135.csv", show_col_types = FALSE)

autopsy_breakdown <- autopsy_breakdown %>%
  left_join(lookup113, by = c("level" = "code"))

# ── summary by sex ─────────────────────────────────────────────
sex_counts <- mort %>%
  filter(autopsy_flag %in% c("Yes", "No")) %>%  
  group_by(sex_lbl) %>%
  summarise(
    autopsy_yes = sum(autopsy_flag == "Yes"),
    total       = n(),
    .groups = "drop"
  )

sex_stat <- sex_counts %>%
  rowwise() %>%
  mutate(
    prop = autopsy_yes / total,
  ) %>%

print(sex_stat)

write_csv(autopsy_breakdown,
          "autopsy_breakdown_demographics_2016.csv")  
print(head(autopsy_breakdown, 20))
```

Is the male/female difference simply a result of men dying younger? The rest of it could be the way men are dying? 
```{r}
reg_dat <- mort %>%                               # keep only Yes/No rows
  filter(autopsy_flag %in% c("Yes", "No")) %>%
  mutate(
    autopsy_bin = as.integer(autopsy_flag == "Yes")  # 1 = autopsy performed
  )

# model 1: sex only
m1 <- lm(autopsy_bin ~ sex_lbl, data = reg_dat)

# model 2: add age (continuous, in years)
m2 <- lm(autopsy_bin ~ sex_lbl + age_years, data = reg_dat)

m3 <- lm(autopsy_bin ~  age_years, data = reg_dat)

# quick summaries
summary(m1)          # baseline difference ⟶ coefficient for sex_lblMale
summary(m2)          # see how that coefficient shrinks once age is added
summary(m3)
anova(m3, m2)

# if you want a tidy comparison table:
bind_rows(
  tidy(m1) %>% mutate(model = "Sex only"),
  tidy(m2) %>% mutate(model = "Sex + age"),
  tidy(m3) %>% mutate(model = "Age only")
)
```

Make plots for autopsy data!
```{r}

autopsy_breakdown <- read_csv(
  "/Users/amymann/Documents/Data Quality Project/data_quality/autopsy_breakdown_demographics_2016.csv",
  show_col_types = FALSE
)

age12_labels <- setNames(
  c("<1", "1–4", "5–14", "15–24", "25–34", "35–44",
    "45–54", "55–64", "65–74", "75–84", "85+", "NS"),
  sprintf("%02d", 1:12)
)

# ── 3. Helper to build one stacked-bar plot ───────────────────────────────────
plot_autopsy_prop <- function(df, var){
  plot_df <- df %>%
    filter(variable == var, autopsy_flag != "Unknown") %>%
    mutate(level = case_when(
      var == "age_grp12"              ~ factor(level,
                                               levels = names(age12_labels),
                                               labels = age12_labels),
      var %in% c("race_lbl", "hisp_lbl") ~ fct_relevel(level),  # keep file order
      TRUE                              ~ fct_infreq(level)     # order by freq
    ))

  ggplot(plot_df, aes(level, prop, fill = autopsy_flag)) +
    geom_col(width = 0.9, position = "stack") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_brewer(palette = "Set1") +
    labs(x = NULL, y = "Proportion of deaths",
         fill = "Autopsy performed",
         title = paste("Autopsy frequency by", var)) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title  = element_text(size = 12, face = "bold"))
}

# ── pick variables for grid  ──────────────────
grid_vars <- c("sex_lbl", "age_grp12", "race_lbl", "hisp_lbl",
               "edu_final", "marital", "inj_work")

# ── make grid plot  ──────────────────────────────────────────────────
grid_plot <- wrap_plots(
  lapply(grid_vars, \(v) plot_autopsy_prop(autopsy_breakdown, v)),
  ncol = 2
) +
  plot_annotation(
    title = "Autopsy proportions across demographic groups (US deaths, 2016)",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

print(grid_plot)
ggsave("autopsy_demographics_grid_2016.png",
       grid_plot, width = 14, height = 10, dpi = 300)

# ── make 113 cause groups plot ───────────────────────────
cause_plot <- autopsy_breakdown %>%
  filter(variable == "cause113",
         autopsy_flag != "Unknown",
         as.numeric(level) >= 1,
         as.numeric(level) <= 113) %>%
  mutate(desc = fct_reorder(level, prop)) %>%   # use plain-English label
  ggplot(aes(x = desc, y = prop, fill = autopsy_flag)) +
    geom_col(position = "stack", width = 0.9) +
    scale_y_continuous(labels = percent_format()) +
    scale_fill_manual(values = c("Yes"="#377eb8",
                                 "No" ="#e41a1c")) +
    labs(x = "Underlying-cause (113-group recode)",
         y = "Proportion of deaths",
         fill = "Autopsy",
         title = "Autopsy frequency by underlying cause, 2016") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4),
          plot.title  = element_text(size = 13, face = "bold"))

print(cause_plot)
ggsave("autopsy_by_cause113_2016.png",
       cause_plot, width = 16, height = 6, dpi = 300)

```

Average number of secondary factors listed 
```{r}
demo_vars <- c("race", "racer5", "educ2003", "sex", "ager52", "marstat", "ucr39")

mort <- mortality |>            
  mutate(extra_factor_count =
           rowSums(across(starts_with("record_"),
                          ~ .x != "" & !is.na(.x)))-1)
extra_factor_summary <- bind_rows(
  lapply(demo_vars, function(v)
    mort %>% 
      group_by(.data[[v]], .drop = FALSE) %>% 
      summarise(
        n_deaths   = n(),
        mean_extra = mean(extra_factor_count, na.rm = TRUE),
        .groups    = "drop"
      ) %>% 
      transmute(
        variable   = v,
        level      = as.character(.data[[v]]),    # ensure character
        n_deaths,
        mean_extra
      )
  )
)


print(extra_factor_summary)
write_csv(extra_factor_summary, "avg_secondary_factors_by_demo.csv")

top10_ucr39 <- mort %>%
  count(ucr39, sort = TRUE, name = "n_deaths") %>%
  slice_head(n = 10) %>%
  left_join(
    mort %>% group_by(ucr39) %>%
      summarise(mean_extra = mean(extra_factor_count, na.rm = TRUE),
                .groups = "drop"),
    by = "ucr39"
  )
print(top10_ucr39)


hist_plot <- ggplot(mort, aes(extra_factor_count)) +
  geom_histogram(binwidth = 1, boundary = -0.5, fill = "#2e86c1") +
  scale_x_continuous(breaks = 0:20) +
  labs(x = "Number of record-axis conditions",
       y = "Deaths",
       title = "Distribution of additional conditions per death") +
  theme_minimal(base_size = 11)

ggsave("hist_extra_factors.png", hist_plot, width = 7, height = 4, dpi = 300)


age_var <- if ("age52" %in% names(mort)) "age52" else "ager52"

age_means <- mort %>%
  group_by(.data[[age_var]]) %>%
  summarise(mean_extra = mean(extra_factor_count, na.rm = TRUE),
            .groups = "drop")

age_plot <- ggplot(age_means,
                   aes(factor(.data[[age_var]]), mean_extra)) +
  geom_col(fill = "#e41a1c") +
  scale_y_continuous(labels = number_format(accuracy = 0.1)) +
  labs(x = "Age category (Recode 52)",
       y = "Mean number of record-axis conditions",
       title = "Average additional conditions by age group") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave("mean_extra_factors_by_age.png", age_plot, width = 8, height = 4, dpi = 300)

```

Frequency of 0 secondary factors listed 
```{r}
demo_vars <- c("racer3", "racer5", "educ2003", "sex", "ager52", "marstat")

mort <- mortality |>            
  mutate(extra_factor_count =
           rowSums(across(starts_with("record_"),
                          ~ .x != "" & !is.na(.x)))-1)
# ── 1. proportion with *zero* secondary factors, by demographic ───────────────
no_sec_summary <- bind_rows(
  lapply(demo_vars, function(v)
    mort %>% 
      group_by(.data[[v]], .drop = FALSE) %>% 
      summarise(
        n_deaths   = n(),
        prop_zero  = mean(extra_factor_count == 0, na.rm = TRUE),
        .groups    = "drop"
      ) %>% 
      transmute(
        variable = v,
        level    = as.character(.data[[v]]),
        n_deaths,
        prop_zero
      )
  )
)

write_csv(no_sec_summary, "prop_no_secondary_by_demo.csv")
print(no_sec_summary)

# ── 2. top-10 UCR-39 causes – prop with no secondary factor ───────────────────
top10_ucr39_none <- mort %>%
  count(ucr39, sort = TRUE, name = "n_deaths") %>%
  slice_head(n = 10) %>%
  left_join(
    mort %>% group_by(ucr39) %>%
      summarise(prop_zero = mean(extra_factor_count == 0, na.rm = TRUE),
                .groups   = "drop"),
    by = "ucr39"
  )

write_csv(top10_ucr39_none, "prop_no_secondary_top10_ucr39.csv")
print(top10_ucr39_none)

# ── 3. bar chart of prop-zero by age52 (or ager52) ────────────────────────────
age_var <- if ("age52" %in% names(mort)) "age52" else "ager52"

age_prop_zero <- mort %>%
  group_by(.data[[age_var]]) %>%
  summarise(prop_zero = mean(extra_factor_count == 0, na.rm = TRUE),
            .groups   = "drop")

ggplot(age_prop_zero, aes(factor(.data[[age_var]]), prop_zero)) +
  geom_col(fill = "#e41a1c") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Age category (Recode 52)",
       y = "Share of deaths with NO secondary factor",
       title = "Percentage with no record-axis conditions by age group") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave("prop_no_secondary_by_age.png", width = 8, height = 4, dpi = 300)

```

Can sex differences in frequency of 0 or 1 secondary factors listed be explained by cause of death and/or age of death?
```{r}
mort <- mortality %>% 
  mutate(
    has_extra_factor = 1- as.integer(                       # 1 = none, 0 = ≥1
      rowSums(across(starts_with("record_"),
                     ~ .x != "" & !is.na(.x))) > 1
    )
  )
mort <- mort %>%                    
  mutate(
    ucr39 = factor(ucr39, levels = 1:39) 
  )

m1 <- lm(has_extra_factor ~ sex, data = mort)

library(MASS)

full_mod  <- lm(has_extra_factor ~ sex + age + ucr39, data = mort)

step_mod  <- stepAIC(full_mod,
                     scope = list(upper = ~ age + sex + ucr39,
                                  lower = ~ 1),   # start at intercept-only
                     direction = "both",
                     trace = FALSE)

summary(step_mod)


# quick summaries
summary(m1)          # baseline difference ⟶ coefficient for sex_lblMale
```

How correlated is everything?

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

